---
title: "Using anatembea to Reconstruct Transmission Patterns: A Vignette"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using anatembea to Reconstruct Transmission Patterns: A Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "images/"
)
# Skip slow model-fitting code during R CMD check
eval_model <- identical(Sys.getenv("NOT_CRAN"), "true")
```

In this vignette, we demonstrate how to use *anatembea* to reconstruct patterns of transmission intensity, seasonality, and malaria burden from monthly malaria prevalence data in children under 5 years old. Later vignettes will demonstrate applying this framework to settings where pregnant women are screened for malaria (using either rapid diagnostic tests or slide microscopy) with data available at a monthly cadence as well as exploring scenarios involving different data cadences or stratifications (e.g., by gravidity).

## Understanding our approach

*anatembea* is a tool designed to integrate [*dust*](https://mrc-ide.github.io/dust/), [*odin*](https://mrc-ide.github.io/odin/), and [*mcstate*](https://mrc-ide.github.io/mcstate) frameworks for fitting a semi-stochastic mechanistic disease transmission model to continuously collected prevalence data using particle Markov chain Monte Carlo (pMCMC). This package specifically fits a semi-stochastic version of the [*malariasimulation*](https://mrc-ide.github.io/malariasimulation/) model to monthly malaria prevalence. While this package was designed for analysis of malaria prevalence measured among pregnant women at their first antenatal care (ANC1) visit, it can also be fit to monthly malaria prevalence in the general population. This vignette walks through a simple example fitting to simulated data from children under 5 years old and outlines various configurable options and settings.

In the [deterministic version](https://mrc-ide.github.io/deterministic-malaria-model/index) of *malariasimulation*, seasonal variation in malaria transmission intensity is governed by a Fourier series, with coefficients pre-estimated from historical rainfall data. While this method provides reasonable seasonal profiles for fitting cross-sectional data collected at multi-year intervals (such as data from Demographic Health Surveys or Malaria Indicator Surveys), it lacks the flexibility required for fitting models to monthly prevalence data collected through ANC1.

To enhance flexibility, we replaced the deterministic seasonal mechanism with a stochastic process. Specifically, mosquito emergence rates now vary according to a random-walk process, where the monthly change in mosquito emergence rate is modelled as: Emergence rate change = *δ*∼*N*(0,*σ*).

Here, *σ* is a volatility constant, and *β*, the mosquito emergence rate, is constrained by an upper limit *βmax* to maintain realism. The pMCMC filters likely emergence rate trajectories by comparing the observed prevalence with the model's estimated prevalence based on these trajectories. The output is a posterior distribution of plausible model trajectories.

Since the relationship between mosquito emergence rates and infection prevalence is encoded within the mechanistic model, *anatembea* allows us to infer trends in key indicators like the entomological inoculation rate (EIR) and clinical incidence, which are otherwise challenging to measure directly.

## Generating a test dataset

At its core, our approach leverages the established relationships between temporal infection prevalence dynamics, transmission, and clinical burden, as defined in the *malariasimulation* model. Consequently, we can validate the inferential framework by generating simulated datasets from the model and verifying that *anatembea* correctly recaptures the underlying seasonal dynamics.

For this vignette, we use a simulation based on rainfall patterns in Tanga, Tanzania, where rainfall follows a characteristic bimodal pattern. This includes:

1.  A short rainy season (March–May)
2.  A longer rainy season (November–January)
3.  An extended dry season (June–October)

Bimodal rainfall patterns, such as those in Tanga, often present a challenging inferential problem for many algorithms attempting to estimate seasonality. However, by applying *anatembea*, we aim to recapture these complex dynamics without supplying the model with prior information about the simulation parameters.

The resulting relationship between malaria prevalence in children under 5, clinical incidence, and EIR in Tanga, Tanzania is illustrated below:

```{r}
library(dplyr)
library(ggplot2)
library(anatembea)
data_target <- anatembea::sim_data_tanzania%>%
  dplyr::filter(month<=zoo::as.yearmon('Dec 2019'))

prev_plot_target <- ggplot(data=data_target)+
  geom_line(aes(x=month,y=prev_05_target),color="#1F78B4")+
  scale_y_continuous(limits=c(0,0.5),expand=c(0,0))+
  labs(y='RDT Prevalence,\nChildren < 5 years')+
  theme_minimal()+
  theme(axis.title.x = element_blank())

inc_plot_target <- ggplot(data=data_target)+
  geom_line(aes(x=month,y=clininc_all_target),color="#33A02C")+
  scale_y_continuous(limits=c(0,NA),expand=c(0,0))+
  labs(y='Clinical Incidence,\nall ages')+
  theme_minimal()+
  theme(axis.title.x = element_blank())

eir_plot_target <- ggplot(data=data_target)+
  geom_line(aes(x=month,y=EIR_target),color="#E31A1C")+
  scale_y_log10(expand=c(0,0))+
  labs(y='EIR')+
  theme_minimal()+
  theme(axis.title.x = element_blank())

library(cowplot)
combined_ggplot_target <- cowplot::plot_grid(prev_plot_target,inc_plot_target,eir_plot_target,ncol=1)



```

```{r}
combined_ggplot_target
```

## Input Data

The only user input required for running *anatembea* is a data frame of monthly prevalence with at least three columns:

1.  `month`: Month (here formatted as 'yearmon', provided by the package 'zoo')
2.  `tested`: The number of individuals tested
3.  `positive`: Of those tested, the number of positive test results

In our example, the data frame `tanga_data_slim` was generated by sampling binomially from the target infection prevalence of the simulation presented above using a monthly sample size of 1000 children.

```{r data}
tanga_data_slim <- anatembea::sim_data_tanzania%>%
  dplyr::select(month,positive,tested)%>%
  dplyr::filter(month<=zoo::as.yearmon('Dec 2019'))
```

```{r, fig.show='hold'}
head(tanga_data_slim, 5)
```

## Running a pMCMC fitting

The `run_pmcmc` function is the central tool to fit our model to observed data. It formats the provided data set, sets up required functions and parameters, and finally runs the pMCMC. Broadly, the flow of actions is as follows:

1.  Process provided data set, specifically formatting necessary time variables.
2.  Format and declare parameters. These include both constant or known parameter values needed for malariasimulation as well as pMCMC parameters that will be fit.
3.  Set particle filter and pMCMC settings.
4.  Run the pMCMC.
5.  Format output.

By default, in the absence of other contextual information, we use the first year’s prevalence (total number of positives in the first year divided by total number of tested in the first year) to calibrate initial pre-observation transmission and immunity, minimising potential seasonal biases of a shorter calibration window. However, *anatembea* provides functionality for the user to adjust this assumption. In the example below, we will assume that, on average before data collection, prevalence in children under 5 was 0.4, as specified in the option `target_prev`.

Other optional user parameters include pMCMC particles/step/chains, the maximum mosquito emergence rate, and treated fraction of clinical cases. No rainfall or other environmental data are required. For available options, see package [documentation](https://mrc-ide.github.io/anatembea/reference/run_pmcmc.html "run_pmcmc documentation").

```{r pmcmc, fig.show='hold',  dependson="data", eval=identical(Sys.getenv("NOT_CRAN"), "true")}
result <- anatembea::run_pmcmc(data_raw=tanga_data_slim,
                               target_prev = 0.4,
                               n_particles=100,
                               n_steps = 10)
```

## Visualising output

We can then visualise how well the pMCMC fit to the prevalence data, as well as unobserved transmission metrics like clinical incidence in the entire population and the entomological inoculation rate (EIR).

First we'll perform some simple diagnostics, including visualising the MCMC trace and calculating the acceptance rate and effective sample size.

```{r diagnostics, eval=identical(Sys.getenv("NOT_CRAN"), "true")}
measure_names <- c('prev_05','clininc_all','EIR','betaa')
months <- tanga_data_slim$month

timelength <- length(months)
num_months <- length(result$history[1,1,])
start_month_index <- (num_months-timelength+1)

mcmc <- result$mcmc[,]
ar <- 1 - coda::rejectionRate(coda::as.mcmc(mcmc))
ess <- coda::effectiveSize(coda::as.mcmc(mcmc))

pars_list <- names(mcmc)
trace_plots <- lapply(pars_list, function(x){
  bayesplot::mcmc_trace(mcmc,pars = x) +
    ggtitle(paste0(x,' / AR: ',round(ar[x],3),' / ESS: ',round(ess[x],1)))+
    theme(title = element_text(size=6),
          axis.title.y = element_blank())
  })
trace_plots
```

In the example above, our diagnostics don't look great. We want an acceptance rate around 20% and an ESS of at least 200 (the higher the better). However, based on some simplified options we've given, we are unlikely to achieve good performance. Besides the data (`data_raw`) and `target_prev`, when we ran the pMCMC, we also specified two additional options, which set the number of particles (`n_particles`) and the number of MCMC iterations (`n_steps`) for the pMCMC algorithm. We gave lower numbers than advised for a real analysis to avoid lengthy computational times for this example. A good starting point for an initial run is about 200 particles and at least 1000 steps (although for a final analysis, the number of steps may be above 10,000 to ensure MCMC convergence and proper mixing).

Even though our MCMC doesn't look great, let's still visualize the model fit to the data:

```{r output_process, eval=identical(Sys.getenv("NOT_CRAN"), "true")}
history.dfs <- lapply(measure_names,function(x){
  as.data.frame(t(result$history[x,,start_month_index:num_months]))%>%
        dplyr::mutate(month=months)%>%
    reshape2::melt(id='month')%>%
    dplyr::group_by(month)%>%
    dplyr::summarise(median=median(value),
              upper=quantile(value,probs=0.975),
              lower=quantile(value,probs=0.025))%>%
    dplyr::mutate(measure=x)
})
names(history.dfs) <- measure_names

prev_plot <- ggplot(data=history.dfs[['prev_05']])+
  geom_ribbon(aes(x=month,ymin=lower,ymax=upper),fill="#A6CEE3")+
  geom_line(aes(x=month,y=median),color="#1F78B4")+
  geom_point(data=tanga_data_slim,aes(x=month,y=positive/tested),fill='black')+
  scale_y_continuous(limits=c(0,0.5),expand=c(0,0))+
  labs(y='RDT Prevalence,\nChildren < 5 years')+
  theme_minimal()+
  theme(axis.title.x = element_blank())

inc_plot <- ggplot(data=history.dfs[['clininc_all']])+
  geom_ribbon(aes(x=month,ymin=lower,ymax=upper),fill="#B2DF8A")+
  geom_line(aes(x=month,y=median),color="#33A02C")+
  scale_y_continuous(limits=c(0,NA),expand=c(0,0))+
  labs(y='Clinical Incidence,\nall ages')+
  theme_minimal()+
  theme(axis.title.x = element_blank())

eir_plot <- ggplot(data=history.dfs[['EIR']])+
  geom_ribbon(aes(x=month,ymin=lower,ymax=upper),fill="#FB9A99")+
  geom_line(aes(x=month,y=median),color="#E31A1C")+
  scale_y_log10(expand=c(0,0))+
  labs(y='EIR')+
  theme_minimal()+
  theme(axis.title.x = element_blank())

combined_ggplot <- cowplot::plot_grid(prev_plot,inc_plot,eir_plot,ncol=1)
# combined_ggplot <- prev_plot+inc_plot+eir_plot+plot_layout(ncol=1)
result$output <- combined_ggplot

```

```{r, fig.show='hold', eval=identical(Sys.getenv("NOT_CRAN"), "true")}
result$output

```
