---
title: "Preparing data and configuration for anatembea"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing data and configuration for anatembea}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "images/preparing-"
)
```

Particle Markov chain Monte Carlo (pMCMC) can be computationally expensive. A
longer run will not repair invalid counts, incorrectly ordered dates, or a
poorly specified initialization. This article therefore prepares and freezes
the analysis specification *before* tuning begins.

At the end of this article you will have:

1. a checked monthly prevalence dataset;
2. a documented model specification; and
3. enough provenance to reproduce the analysis.

The next article, [Tuning an anatembea
pMCMC](tuning-pmcmc.html), uses that frozen specification to select a particle
count and proposal covariance.

## Required data

For a single comparison group, `run_pmcmc()` expects one row per month and at
least these columns:

- `month`: a monthly time variable, represented here by `zoo::yearmon`;
- `tested`: the number of people tested; and
- `positive`: the number who tested positive.

The package includes simulated data that we can use throughout this series.

```{r example-data}
analysis_data <- anatembea::sim_data_tanzania |>
  dplyr::select(month, positive, tested) |>
  dplyr::filter(month <= zoo::as.yearmon("Dec 2019"))

head(analysis_data, 4)
```

For paired ANC comparisons such as `comparison = "pgmg"`, prepare separate
primigravidae and multigravidae datasets with the same three columns and pass
them as `data_raw_pg` and `data_raw_mg`. See
[`run_pmcmc()`](../reference/run_pmcmc.html) for all supported comparison
groups.

## Validate before fitting

The helper below checks structural errors while allowing a month in which both
counts are missing. `anatembea` can treat such a month as an unobserved time
point, but a missing `positive` paired with a non-missing `tested` (or vice
versa) is ambiguous and should be resolved.

```{r validation-helper}
validate_monthly_prevalence <- function(data) {
  required <- c("month", "positive", "tested")
  missing_columns <- setdiff(required, names(data))
  if (length(missing_columns) > 0) {
    stop("Missing required columns: ", paste(missing_columns, collapse = ", "))
  }

  month_key <- format(data$month, "%Y-%m")
  if (anyNA(month_key)) stop("`month` contains missing or unparseable values.")
  if (is.unsorted(data$month)) stop("Rows must be in chronological order.")
  if (anyDuplicated(month_key)) stop("Each month must appear at most once.")

  one_count_missing <- xor(is.na(data$positive), is.na(data$tested))
  if (any(one_count_missing)) {
    stop("`positive` and `tested` must be missing together.")
  }

  observed <- !is.na(data$positive)
  invalid_counts <- data$positive[observed] < 0 |
    data$tested[observed] < 0 |
    data$positive[observed] > data$tested[observed]
  if (any(invalid_counts)) {
    stop("Counts must satisfy 0 <= positive <= tested.")
  }

  expected <- format(
    seq(min(data$month), max(data$month), by = 1 / 12),
    "%Y-%m"
  )
  gaps <- setdiff(expected, month_key)
  if (length(gaps) > 0) {
    warning("Missing calendar months: ", paste(gaps, collapse = ", "))
  }

  invisible(data)
}

validate_monthly_prevalence(analysis_data)
```

A gap is a month with no row; a missing observation is a row with both counts
set to `NA`. Decide deliberately which representation your analysis uses.
Also plot `positive / tested` and `tested` over time. Abrupt changes may be
real, but they often reveal altered reporting systems, duplicated extracts, or
changes in the population being tested.

```{r data-plot, fig.height=4, fig.width=7}
plot_data <- analysis_data |>
  dplyr::mutate(prevalence = positive / tested)

ggplot2::ggplot(plot_data, ggplot2::aes(month, prevalence)) +
  ggplot2::geom_line(colour = "#1F78B4") +
  ggplot2::geom_point(ggplot2::aes(size = tested), alpha = 0.6) +
  ggplot2::scale_y_continuous(limits = c(0, 1)) +
  ggplot2::labs(x = NULL, y = "Observed prevalence", size = "Tested") +
  ggplot2::theme_minimal()
```

## Freeze the scientific specification

Write down the settings that determine what the model means, not just how
long it runs. At minimum, record:

- `comparison` and the population represented by each input series;
- `initial`, `target_prev`, and the initial-condition window;
- `check_flexibility` and `start_pf_time`;
- treatment and seasonality assumptions, including geographic inputs;
- fitted parameter names and their order; and
- any departure from package defaults.

For example:

```{r specification}
analysis_spec <- list(
  analysis_id = "tanzania-u5-example",
  comparison = "u5",
  initial = "fitted",
  check_flexibility = TRUE,
  start_pf_time = 30 * 12,
  prop_treated = 0.4,
  seasonality_on = TRUE,
  fitted_parameters = c("log_init_EIR", "volatility")
)

analysis_spec
```

### Choose an initialization mode

Use `initial = "fitted"` when initial transmission intensity should be inferred
from the data. In the current implementation, the fitted parameters are
ordered as `log_init_EIR` and then `volatility`. Proposals for `log_init_EIR`
are bounded between `log(1e-3)` and `log(500)` before model evaluation.

Use `initial = "informed"` when an equilibrium EIR or target prevalence is a
scientifically defensible input. Only `volatility` is then sampled. Because
these modes have different parameter spaces, they cannot share a proposal
covariance.

## Record software, inputs, and seeds

Package behavior can change between releases. Save versions and source
revisions alongside every submitted run.

```{r provenance}
provenance <- list(
  recorded_at = format(Sys.time(), tz = "UTC"),
  R = R.version.string,
  packages = vapply(
    c("anatembea", "mcstate", "dust", "odin.dust", "odin"),
    function(x) as.character(utils::packageVersion(x)),
    character(1)
  ),
  git_commit = tryCatch(
    system2("git", c("rev-parse", "HEAD"), stdout = TRUE, stderr = FALSE),
    error = function(e) NA_character_
  ),
  seeds = c(smoke = 1101L, pilot = 2101L, scale_1 = 3101L)
)

provenance
```

For file-based inputs, `tools::md5sum("path/to/input.csv")` provides a compact
way to detect whether the file has changed. A project-specific library managed
with [renv](https://rstudio.github.io/renv/) can record package versions, but
the resolved `anatembea` commit or release should still be saved explicitly.

Do not install a moving branch such as `HEAD` in a production job without
recording the resolved commit.

## Gate: ready to tune?

Proceed only when all of the following are true:

- counts, dates, groups, and gaps have been reviewed;
- initialization and seasonality return finite, plausible states;
- the fitted parameter names and order are recorded;
- model assumptions are frozen; and
- inputs, software versions, and unique seeds can be reconstructed.

You are now ready to [tune the pMCMC](tuning-pmcmc.html).
