---
title: "Tuning an anatembea pMCMC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning an anatembea pMCMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "images/tuning-"
)
```

Tuning answers two practical questions: how many particles are needed for a
stable likelihood estimate, and how should the proposal move through the
parameter space? It is a separate stage from scientific inference.

This article assumes that the data and scientific settings have passed the
checks in [Preparing data and configuration for
anatembea](preparing-an-analysis.html). It ends with a provisional particle
count and proposal covariance. Those settings are tested with independent
chains in [Running and assessing production pMCMC
chains](production-pmcmc.html).

The numerical values below are starting points, not universal thresholds.
Data length, prevalence, missingness, model options, and computing resources
all affect pMCMC performance.

## The tuning sequence

| Stage | Typical starting design | Question answered |
|---|---|---|
| Smoke test | 100 particles, 10--20 iterations | Does the complete execution path work? |
| Particle check | Repeated filters at 50--400 particles | Is the likelihood stable enough? |
| Covariance pilot | 500 iterations, one chain | Which directions should the proposal move? |
| Scale sweep | Three 500-iteration trials | How far should it move? |

Do not use smoke-test or tuning draws for final inference.

## 1. Build the starting proposal

A proposal covariance describes both the size and direction of proposed MCMC
moves. For informed initialization, only `volatility` is fitted, so the
identity covariance is one-dimensional:

```{r identity-one}
proposal_informed <- diag(1)
dimnames(proposal_informed) <- list("volatility", "volatility")
proposal_informed
```

For fitted initialization, the order is `log_init_EIR`, then `volatility`:

```{r identity-two}
parameter_names <- c("log_init_EIR", "volatility")
proposal_fitted <- diag(length(parameter_names))
dimnames(proposal_fitted) <- list(parameter_names, parameter_names)
proposal_fitted
```

The diagonal entries are proposal variances. Off-diagonal entries allow the
parameters to move together. An identity matrix is a neutral way to test the
code path, but it usually ignores the posterior's scale and correlation and
is therefore not a production proposal.

Always check the matrix before starting a costly run:

```{r covariance-check}
check_covariance <- function(x, parameter_names) {
  stopifnot(
    is.matrix(x),
    identical(dim(x), rep(length(parameter_names), 2L)),
    isTRUE(all.equal(x, t(x))),
    all(is.finite(x))
  )
  if (min(eigen(x, symmetric = TRUE, only.values = TRUE)$values) <= 0) {
    stop("The proposal covariance must be positive definite.")
  }
  invisible(x)
}

check_covariance(proposal_fitted, parameter_names)
```

## 2. Run a smoke test

The smoke test verifies compilation, arguments, output structure, versions,
and cluster wiring. The example is not evaluated while building the vignette
because model compilation and pMCMC are deliberately expensive.

```{r smoke-test, eval=FALSE}
smoke <- anatembea::run_pmcmc(
  data_raw = analysis_data,
  comparison = "u5",
  initial = "fitted",
  check_flexibility = TRUE,
  proposal_matrix = proposal_fitted,
  n_particles = 100,
  n_steps = 20,
  n_threads = 1,
  n_chains = 1,
  n_workers = 1,
  seed = 1101L,
  output_level = "minimal",
  save_state = FALSE,
  save_trajectories = FALSE
)

names(smoke)
head(smoke$pars)
```

`output_level = "minimal"` retains the likelihood inputs but avoids retaining
unneeded model outputs. With `save_trajectories = FALSE`, `history` and `times`
will be `NULL`. This is appropriate for short tuning runs that only need
parameter and likelihood diagnostics.

A successful smoke test is evidence that the workflow executes—not that the
chain converged or that the proposal is suitable.

## 3. Check the particle count

The particle filter produces a noisy estimate of the likelihood. Too few
particles can cause highly variable likelihood estimates, repeated MCMC
states, or filter failures; too many increase runtime without necessarily
improving efficiency.

At a representative fixed parameter vector, repeat the particle filter with
independent seeds at particle counts such as 50, 100, 200, and 400. Record:

- the variance or standard deviation of the estimated log likelihood;
- particle-filter and ODE failures;
- elapsed time and memory; and
- the exact parameter vector and seed.

The current high-level `run_pmcmc()` interface performs a complete pMCMC rather
than exposing a standalone fixed-parameter likelihood replicate. Consequently,
this check requires a project-specific particle-filter wrapper built from the
same model specification. Do not substitute a single short MCMC chain and call
its changing likelihood values likelihood replicates.

A useful results layout is:

```{r particle-table}
particle_check <- data.frame(
  particles = c(50, 100, 200, 400),
  likelihood_sd = NA_real_,
  failures = NA_integer_,
  elapsed_seconds = NA_real_,
  memory_mb = NA_real_
)

particle_check
```

Select the smallest count that provides a sufficiently stable likelihood with
no systematic failures. One hundred particles is often a useful tuning start
and 200 a confirmation or production start, but the empirical check takes
precedence. Hold particle count fixed while comparing proposal scales.

## 4. Estimate a covariance from a pilot

Run a separate pilot for every materially different time series or analysis
unit. A typical starting design is one 500-iteration chain using the selected
tuning particle count and a unique seed.

```{r covariance-pilot, eval=FALSE}
pilot <- anatembea::run_pmcmc(
  data_raw = analysis_data,
  comparison = "u5",
  initial = "fitted",
  check_flexibility = TRUE,
  proposal_matrix = proposal_fitted,
  n_particles = 100,
  n_steps = 500,
  seed = 2101L,
  output_level = "minimal",
  save_state = FALSE,
  save_trajectories = FALSE
)
```

Discard burn-in, retain finite parameter draws, and estimate their covariance:

```{r estimate-covariance, eval=FALSE}
burn_in <- floor(nrow(pilot$pars) / 2)
retained <- pilot$pars[(burn_in + 1):nrow(pilot$pars), parameter_names,
                       drop = FALSE]
retained <- retained[stats::complete.cases(retained), , drop = FALSE]

if (nrow(retained) < 50) {
  stop("Too few usable pilot draws; diagnose or extend the pilot.")
}

pilot_covariance <- stats::cov(retained)
pilot_covariance <- (pilot_covariance + t(pilot_covariance)) / 2

# Correct only small numerical defects; do not hide an unidentified parameter.
eigen_floor <- 1e-8
decomp <- eigen(pilot_covariance, symmetric = TRUE)
decomp$values <- pmax(decomp$values, eigen_floor)
pilot_covariance <- decomp$vectors %*%
  diag(decomp$values) %*% t(decomp$vectors)
dimnames(pilot_covariance) <- list(parameter_names, parameter_names)

check_covariance(pilot_covariance, parameter_names)
pilot_covariance
```

Review trace plots, parameter ranges, acceptance, repeated states, failures,
and correlation before using this matrix. If the pilot barely moves or yields
fewer than about 50 usable retained draws, do not conceal the problem with
aggressive regularization. Diagnose it or repeat a longer pilot.

## 5. Sweep the proposal scale

Use the pilot acceptance rate to choose an initial scale grid:

| Pilot acceptance | Covariance multipliers |
|---|---|
| Below 0.35 or unavailable | 0.75, 1.0, 1.5 |
| 0.35--0.45 | 1.0, 1.5, 2.0 |
| Above 0.45 | 1.5, 2.0, 2.83 |

The multiplier here is applied directly to the covariance. If you instead
intend to multiply proposal standard deviations, square that multiplier before
applying it to the covariance.

```{r scale-grid, eval=FALSE}
pilot_mcmc <- coda::as.mcmc(pilot$pars[, parameter_names, drop = FALSE])
pilot_acceptance <- 1 - coda::rejectionRate(pilot_mcmc)
pilot_acceptance <- min(pilot_acceptance)

scales <- if (pilot_acceptance < 0.35) {
  c(0.75, 1, 1.5)
} else if (pilot_acceptance <= 0.45) {
  c(1, 1.5, 2)
} else {
  c(1.5, 2, 2.83)
}

scale_runs <- Map(function(multiplier, seed) {
  anatembea::run_pmcmc(
    data_raw = analysis_data,
    comparison = "u5",
    initial = "fitted",
    check_flexibility = TRUE,
    proposal_matrix = multiplier * pilot_covariance,
    n_particles = 100,
    n_steps = 500,
    seed = seed,
    output_level = "minimal",
    save_state = FALSE,
    save_trajectories = FALSE
  )
}, scales, c(3101L, 3102L, 3103L))
```

Summarize every parameter, not just the best-looking one:

```{r summarize-scales, eval=FALSE}
summarize_run <- function(result, scale) {
  chain <- coda::as.mcmc(result$pars[, parameter_names, drop = FALSE])
  acceptance <- 1 - coda::rejectionRate(chain)
  ess <- coda::effectiveSize(chain)
  hours <- as.numeric(result$run_time, units = "hours")

  data.frame(
    scale = scale,
    parameter = names(ess),
    acceptance = acceptance[names(ess)],
    ess = as.numeric(ess),
    ess_per_hour = as.numeric(ess) / hours,
    repeated_proportion = 1 - acceptance[names(ess)]
  )
}

scale_summary <- do.call(
  rbind,
  Map(summarize_run, scale_runs, scales)
)
scale_summary
```

Rank successful trials primarily by their minimum ESS per hour across fitted
parameters. Acceptance around 0.25--0.45 is a useful guide for this
low-dimensional random-walk proposal, but trace movement, stable trajectories,
and absence of failures matter more than hitting a particular number. R-hat is
not informative for a single short chain.

If every scale performs poorly, make one targeted adjustment: improve the
covariance estimate, alter the grid, increase particles if the likelihood is
noisy, or investigate model stiffness. Repeatedly increasing chain length
without a diagnosis is not tuning.

## Gate: ready for confirmation?

Freeze a provisional configuration only when you have recorded:

- the particle-count evidence;
- the finite, symmetric, positive-definite covariance and parameter order;
- the selected covariance multiplier;
- acceptance, repeated states, ESS/hour, traces, and failures for every trial;
- all seeds and runtimes; and
- a reason for selecting the winning configuration.

Next, test that configuration with [independent confirmation
chains](production-pmcmc.html).
